---
title: "BMIN5030 Final Project: Diabetes Prediction"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Introduction

Diabetes is a growing public health concern globally, affecting millions of individuals and contributing to a significant burden on healthcare systems. Early detection and management of diabetes are crucial to mitigating its complications. This project explores the use of predictive modeling techniques to identify individuals at risk of diabetes using the Behavioral Risk Factor Surveillance System (BRFSS) 2015 dataset.

The dataset includes health indicators collected through surveys, offering insights into the behavioral, demographic, and clinical characteristics of respondents. By leveraging machine learning algorithms, this study aims to achieve the following objectives:

1.  Predict the likelihood of diabetes based on survey responses.

2.  Identify the most significant risk factors contributing to diabetes.

3.  Evaluate and compare the performance of various machine learning models, focusing on metrics such as accuracy, ROC AUC, and Matthews Correlation Coefficient (MCC).

The workflow combines data preprocessing, feature selection, exploratory data analysis, and model development using LightGBM. This document provides a step-by-step guide to the analytical process, from data cleaning to model evaluation.

# Data Preprocessing

```{r}
# Load required libraries
library(readxl)
library(tidyverse)
library(FNN)

# Load the dataset
diabetesdata <- read_excel("diabetes_012_health_indicators_BRFSS2015.xlsx")

# Filter out rows with Diabetes_012 == 1 and rename the column
diabetesdata <- diabetesdata %>%
  filter(Diabetes_012 != 1) %>%
  rename(Diabetes_binary = Diabetes_012)

# Convert target into a factor
diabetesdata$Diabetes_binary <- as.factor(diabetesdata$Diabetes_binary)

```

## Class Imbalance

```{r}
diabetesdata %>%
  ggplot(aes(x = Diabetes_binary)) +
  geom_bar() +
  labs(
    x = "Diabetes_binary",
    y = "Count"
  ) +
  theme_minimal()
```

We have a significant class imbalance. Since the dataset is quiet large, down sampling the 'no diabetes' class is a viable approach for addressing the class imbalance.

```{r}
# Down-sample
data_balanced <- diabetesdata %>%
  group_by(Diabetes_binary) %>%
  sample_n(size = min(table(diabetesdata$Diabetes_binary))) %>%
  ungroup()


```

```{r}
data_balanced %>%
  ggplot(aes(x = Diabetes_binary)) +
  geom_bar() +
  labs(
    x = "Diabetes_binary",
    y = "Count"
  ) +
  theme_minimal()
```

# Exploratory Data Analysis

Let's start with broad overview of the dataset using the skimr package.

```{r}
library(skimr)
skim(data_balanced)
```

We have 21 numeric predictors in the dataset with no missing values. Let's check how many are binary predictors.

```{r}
# Count the number of binary variables in the dataset
binary_variable_count <- sum(sapply(diabetesdata, function(column) {
  is.numeric(column) && all(column %in% c(0, 1))
}))

cat("Number of binary variables:", binary_variable_count)

```

## Correlations

Now let's check the spearman correlations between the predictors.

```{r}
library(reshape2)

# Compute the correlation matrix
correlation_matrix <- cor(data_balanced[,-1], use = "complete.obs",method = "spearman")

# Melt the correlation matrix for ggplot
melted_corr <- melt(correlation_matrix)

# Create the heat map
ggplot(melted_corr, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Heat Map",
       x = "Features",
       y = "Features") +
  coord_fixed()
```

```{r}

# Melt the correlation matrix
melted_corr <- melt(correlation_matrix)

# Remove self-correlations
melted_corr <- melted_corr[melted_corr$Var1 != melted_corr$Var2, ]

# Sort by absolute correlation values
melted_corr <- melted_corr %>%
  arrange(desc(abs(value)))

# Get the top 5 positive correlations
top_positive <- melted_corr %>%
  filter(value > 0) %>%
  head(5)

# Get the top 5 negative correlations
top_negative <- melted_corr %>%
  filter(value < 0) %>%
  head(5)

cat("Top 5 Positive Correlations:\n")
print(top_positive)

cat("\nTop 5 Negative Correlations:\n")
print(top_negative)

```

As we might expect physical health, general health, and difficulty walking all have a fairly high correlation with one another.

## Diabetes Comparison Plots

Now let's visualize how some of these predictors relate to the target variable, Diabetes_binary.

```{r}
# Relabel Diabetes_binary levels
data_balanced$Diabetes_binary <- factor(
  data_balanced$Diabetes_binary,
  levels = c(0, 2),
  labels = c("No Diabetes", "Diabetes")
)

# Calculate proportions within each GenHlth level
genhlth_proportions <- data_balanced %>%
  group_by(GenHlth, Diabetes_binary) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(GenHlth) %>%
  mutate(Proportion = Count / sum(Count))

# Plot proportions for GenHlth 
ggplot(genhlth_proportions, aes(
  x = factor(GenHlth, levels = 5:1,  
             labels = c("Poor", "Fair", "Good", "Very Good", "Excellent")),
  y = Proportion, fill = Diabetes_binary)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
    x = "General Health (GenHlth)",
    y = "Proportion",
    fill = "Diabetes Status"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# Calculate proportions within each Education level
education_proportions <- data_balanced %>%
  group_by(Education, Diabetes_binary) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(Education) %>%
  mutate(Proportion = Count / sum(Count))

# Define labels
education_labels <- c(
  "Never Attended School", "Elementary", "Junior High", "Senior High",
  "Undergraduate", "Graduate"
)

# Plot proportions for Education
ggplot(education_proportions, aes(x = factor(Education, levels = 1:6, labels = education_labels),
                                  y = Proportion, fill = Diabetes_binary)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
       x = "Education Level",
       y = "Proportion",
       fill = "Diabetes_binary") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# Calculate proportions within each HighBP level
highbp_proportions <- data_balanced %>%
  group_by(HighBP, Diabetes_binary) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(HighBP) %>%
  mutate(Proportion = Count / sum(Count))

# Plot proportions for HighBP
ggplot(highbp_proportions, aes(x = factor(HighBP, levels = 0:1, labels = c("No", "Yes")),
                               y = Proportion, fill = Diabetes_binary)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
       x = "High Blood Pressure (HighBP)",
       y = "Proportion",
       fill = "Diabetes_binary") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

```

```{r}
# Calculate proportions within each Age group
age_proportions <- data_balanced %>%
  group_by(Age, Diabetes_binary) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(Age) %>%
  mutate(Proportion = Count / sum(Count))

# Define labels
age_labels <- c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", 
                "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+")

# Plot proportions for Age
ggplot(age_proportions, aes(x = factor(Age, levels = 1:13, labels = age_labels),
                            y = Proportion, fill = Diabetes_binary)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
       x = "Age Group",
       y = "Proportion",
       fill = "Diabetes_binary") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# Calculate proportions within each CholCheck level
cholcheck_proportions <- data_balanced %>%
  group_by(CholCheck, Diabetes_binary) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(CholCheck) %>%
  mutate(Proportion = Count / sum(Count))

# Define labels
cholcheck_labels <- c("No", "Yes")

# Plot proportions for CholCheck
ggplot(cholcheck_proportions, aes(x = factor(CholCheck, levels = 0:1, labels = cholcheck_labels),
                                  y = Proportion, fill = Diabetes_binary)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
    x = "Cholesterol Check Status",
    y = "Proportion",
    fill = "Diabetes Status"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

```{r}
# Calculate proportions within each Income group
income_proportions <- data_balanced %>%
  group_by(Income, Diabetes_binary) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(Income) %>%
  mutate(Proportion = Count / sum(Count))

# Define labels
income_labels <- c(
  "< $10k", "$10k-$15k", "$15k-$20k", "$20k-$25k",
  "$25k-$35k", "$35k-$50k", "$50k-$75k", "≥ $75k"
)

# Plot proportions for Income
ggplot(income_proportions, aes(x = factor(Income, levels = 1:8, labels = income_labels),
                               y = Proportion, fill = Diabetes_binary)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
       x = "Income Level",
       y = "Proportion",
       fill = "Diabetes_binary") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}

# Create the count plot
ggplot(data_balanced, aes(x = BMI, fill = Diabetes_binary)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("blue", "red")) +
  labs(
    x = "BMI",
    y = "Count",
    fill = "Diabetes Status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

```

There appear to be significant differences between the diabetes and no diabetes groups based on these predictors. In particular, general health appears to have a strong correlation with diabetes.

## Unsupervised Analysis

```{r}
library(umap)


# Take a random sample of 500 cases for each group
data_subsample <- data_balanced %>%
  group_by(Diabetes_binary) %>%
  sample_n(size = 500) %>%
  ungroup()

# Extract Diabetes_binary for coloring
diabetes_binary <- data_subsample$Diabetes_binary

# Remove Diabetes_binary from the dataset for UMAP
umap_data <- data_subsample %>%
  select(-Diabetes_binary)

# Scale the data
scaled_data <- scale(umap_data)

# Perform UMAP
set.seed(42)
umap_results <- umap(scaled_data, n_neighbors = 15, min_dist = 0.1, n_components = 2)

# Extract UMAP embeddings
umap_embeddings <- as.data.frame(umap_results$layout)
colnames(umap_embeddings) <- c("UMAP1", "UMAP2")

# Add Diabetes_binary back to UMAP results
umap_embeddings$Diabetes_binary <- diabetes_binary

# Plot with consistent labels
ggplot(umap_embeddings, aes(x = UMAP1, y = UMAP2, color = Diabetes_binary)) +
  geom_point(alpha = 0.7, size = 2) +  # Points
  stat_ellipse(aes(fill = Diabetes_binary), geom = "polygon", alpha = 0.2, color = NA) +  # Transparent ellipses
  scale_color_manual(values = c("red", "blue")) +
  scale_fill_manual(values = c("red", "blue")) +
  labs(
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Diabetes Status",
    fill = "Diabetes Status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )


```

There doesn't seem to be too much separation between people with and without diabetes. This might be a challenge for a predictive model.

# Feature Selection

Let's try to identify which features are most important for predicting whether a person has diabetes or not.

```{r}
library(gt)
library(gridExtra)
library(grid)


# Split predictors (X) and target (Y)
X <- data_balanced %>% select(-Diabetes_binary)  
Y <- data_balanced$Diabetes_binary

# Perform Chi-squared test for each feature
chi_sq_results <- sapply(X, function(column) {
  chisq.test(table(column, Y))$statistic
})

# Create a data frame with feature scores
f_Scores <- data.frame(
  Feature = names(chi_sq_results),
  Score = chi_sq_results
)

# Sort the scores in descending order
f_Scores <- f_Scores %>% arrange(desc(Score))

# Create a publication-quality table for the top 10 features
top_10_features <- f_Scores %>% head(10)

gt_table <- top_10_features %>%
  gt() %>%
  tab_header(title = "Top 10 Features by Chi-Squared Score") %>%
  fmt_number(columns = vars(Score), decimals = 2) %>%
  cols_label(
    Feature = "Feature Name",
    Score = "Chi-Squared Score"
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  )

grid.table(as.data.frame(top_10_features))

```

It looks like general health and high blood pressure are most associated with diabetes here.

Now let's look at the odds ratios from a logistic regression model.

```{r}
# Load necessary libraries
library(caret)
library(broom)
library(ggplot2)

# Split the data into training and testing sets
set.seed(123)
train_index <- createDataPartition(data_balanced$Diabetes_binary, p = 0.75, list = FALSE)
train_data <- data_balanced[train_index, ]
test_data <- data_balanced[-train_index, ]

# Fit a logistic regression model
logistic_model <- glm(
  Diabetes_binary ~ .,  # Include all predictors
  data = train_data,
  family = binomial
)

# Summarize the model
summary(logistic_model)

coefficients <- tidy(logistic_model, conf.int = TRUE)
coefficients <- coefficients %>%
  mutate(Odds_Ratio = exp(estimate), 
         Lower_CI = exp(conf.low), 
         Upper_CI = exp(conf.high))

print(coefficients)

# Plot odds ratios
ggplot(coefficients, aes(x = reorder(term, Odds_Ratio), y = Odds_Ratio)) +
  geom_point(color = "blue") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.2, color = "blue") +
  coord_flip() +
  labs(
    title = "Odds Ratios and Confidence Intervals",
    x = "Predictors",
    y = "Odds Ratio"
  ) +
  theme_minimal()

# Make predictions on the test set
test_data <- test_data %>%
  ungroup() %>%  
  mutate(
    Predicted_Probability = predict(logistic_model, newdata = test_data, type = "response"),
    Predicted_Class = ifelse(Predicted_Probability > 0.5, "Diabetes", "No Diabetes")
  )

# Evaluate the model using a confusion matrix
confusion <- confusionMatrix(
  as.factor(test_data$Predicted_Class),
  as.factor(test_data$Diabetes_binary)
)
print(confusion)

# Plot ROC Curve
library(pROC)
roc_curve <- roc(test_data$Diabetes_binary, test_data$Predicted_Probability)
plot(roc_curve, col = "blue", main = "ROC Curve")
auc <- auc(roc_curve)
cat("AUC:", auc, "\n")

```

Interesting! The CholCheck variable has the highest odds ratio by far. This variable represents people who have had their cholesterol checked within the past 5 years (yes or no).

Finally, let's utilize the Boruta package to identify important variables.

```{r}
library(Boruta)

set.seed(42)  
boruta_result <- Boruta(
  Diabetes_binary ~ .,  
  data = data_balanced,
  maxRuns = 11, 
  doTrace = 2           
)

# Display the Boruta results
print(boruta_result)

# Plot the importance of variables
plot(boruta_result, las = 2, cex.axis = 0.7)

# Finalize the selection of important variables
final_vars <- getSelectedAttributes(boruta_result, withTentative = TRUE)
cat("Important Variables:\n", final_vars, "\n")

# Display a detailed summary of results
boruta_summary <- attStats(boruta_result)
print(boruta_summary)
```

It looks like general health, BMI, high blood pressure, and age were most important for predicting diabetes. No doctor because of cost was least important and even lower than the shadow max importance provided by Boruta. As a result, we will drop this predictor before training the final predictive model.\
\

```{r}
data_balanced<-data_balanced %>%
  select(-NoDocbcCost)
```

# Predictive Model

Let's train a light gradient boosting machine (LightGBM) using tidymodels. We can tune the hyper parameters using 10-fold cross validation.

```{r}
library(tidymodels)
library(bonsai)
library(future) 
library(lightgbm)

set.seed(123)



# Split the data into training and testing sets
data_split <- initial_split(data_balanced, prop = 3/4, strata = Diabetes_binary)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Create cross-validation folds
cv_fold <- vfold_cv(train_data, v = 10, strata = "Diabetes_binary")

# Create a recipe for preprocessing
rec <- recipe(Diabetes_binary ~ ., data = train_data) %>%
  step_YeoJohnson(all_numeric_predictors())  # Transform numeric features to reduce skew



# LightGBM model specification
lgbm_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  mtry = tune(),
  min_n = tune(),
  loss_reduction = tune()
) %>%
  set_engine("lightgbm",
             is_unbalance = TRUE, 
             num_leaves = tune(),
             nthread = future::availableCores()) %>%
  set_mode("classification")

# Combine recipe and model into a workflow
wf <- workflow() %>%
  add_model(lgbm_spec) %>%
  add_recipe(rec)

# Define tuning parameters
lgbm_params <- wf %>%
  extract_parameter_set_dials() %>%
  update(
    trees = trees(c(300, 1500)),
    mtry = mtry(range = c(5, 21)),
    min_n = min_n(range = c(30, 90)),
    tree_depth = tree_depth(range = c(8, 25)),
    learn_rate = learn_rate(range = c(-4, -1.5)), 
    num_leaves = num_leaves(c(100, 1000))
  ) %>%
  finalize(train_data)

# Control settings for tuning
lgbm_ctrl <- control_grid(
  verbose = TRUE,
  save_pred = TRUE,
  save_workflow = TRUE
)

# Perform hyperparameter tuning
set.seed(123)
lgbm_res <- tune_grid(
  wf,
  resamples = cv_fold,
  grid = 20,
  control = lgbm_ctrl,
  # Choose metrics appropriate for classification
  metrics = metric_set(accuracy, roc_auc, mcc),
  param_info = lgbm_params
)

# View tuning results
autoplot(lgbm_res)

# Select the best hyperparameters by a chosen metric, e.g. "mcc"
best_params <- select_best(lgbm_res, metric = "accuracy")

# Finalize the workflow with the best hyperparameters
final_wf <- finalize_workflow(wf, best_params)

# Fit the final model to the entire training data
set.seed(123)
final_model <- fit(final_wf, data = train_data)

# Generate predictions on the test set
test_predictions <- predict(final_model, new_data = test_data, type = "class") %>%
  bind_cols(predict(final_model, new_data = test_data, type = "prob")) %>%
  bind_cols(test_data)

# Calculate performance metrics
metrics <- yardstick::metric_set(accuracy, mcc, roc_auc)
test_metrics <- metrics(test_predictions, truth = Diabetes_binary, estimate = .pred_class, .pred_Diabetes)
print(test_metrics)

# Confusion matrix
conf_mat <- conf_mat(test_predictions, truth = Diabetes_binary, estimate = .pred_class)
autoplot(conf_mat, type = "heatmap") +
  labs(title = "Confusion Matrix")

# Extract underlying LightGBM model
final_fit <- extract_fit_parsnip(final_model)

# Get feature importance
importance <- lightgbm::lgb.importance(final_fit$fit)

# Convert to a data frame for plotting
importance_df <- as.data.frame(importance)

# Plot feature importance
ggplot(importance_df, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Importance", x = "Predictors", y = "Gain") +
  theme_minimal() +
  theme(text = element_text(size = 14))

```

It looks like this model was barely an improvement from the simple logistic regression model we trained earlier. We can see that general health and high blood pressure were ranked as the most important variables for this model.

# Conclusion

Predicting diabetes from this dataset of survey questions proved to be challenging for the models we trained. Expanding the set of variables to include more demographic information may improve model performance. Overall, it seems like self-rated general health was the best indicator of whether or not a person has diabetes.
